---
title: "Analysis of PPS Data"
output:
  html_document: default
  pdf_document: default
---

We are going to start with processed data provided by Dr. Gregor
Theilmeier available in the provided spreadsheet.

```
## Added ... option to function: installIfNeeded
installIfNeeded <- function(packages, ...) {
    installedPackages <- installed.packages()[, 1]
    toInstall <- setdiff(packages, installedPackages)
    if (length(toInstall) > 0) install.packages(toInstall, ...)
}
installIfNeeded(c("readxl", "dplyr",
                  "cowplot", "ggplot2", "ggfortify",
                  "elasticnet", "glmnet", "knitr"), repos="https://cloud.r-project.org")
```

```{r}
set.seed(1782682)
library(readxl)
library(ggplot2)
library(dplyr)
spreadsheet <- "P1-Biomarker-Brad-Naras-short.xlsx"
d <- read_excel(path = spreadsheet, sheet = 1)
names(d) <- c("ID", "gender", "age", "cyanosis", "procedure", "expECCtime",
              "expXclamp", "PPS", "orgfail", "ARF", "circfail", "lungfail",
              "ventilator", "ICU", "hospital", "mortality", "timeOnCPB",
              "xclamp", "circarrest", "bodyweight", "fluidbal", "relfluidbal",
              "catecholamine", "infection", "antibiotics", "posBC", "IL1apre", "IL1apost",
              "IL1aD", "IL2rapre", "IL2rapost", "IL2raD", "IL3pre", "IL3post",
              "IL3D", "IL12pre", "IL12post", "IL12D", "IL16pre", "IL16post",
              "IL16D", "IL18pre", "IL18post", "IL18D", "CTACKpre", "CTACKpost",
              "CTACKD", "GROapre", "GROapost", "GROaD", "HGFpre", "HGFpost",
              "HGFD", "ICAM1pre", "ICAM1post", "ICAM1D", "LIFpre", "LIFpost",
              "LIFD", "MCP3pre", "MCP3post", "MCP3D", "MCSFpre", "MCSFpost",
              "MCSFD", "MIFpre", "MIFpost", "MIFD", "MIGpre", "MIGpost", "MIGD",
              "bNGFpre", "bNGFpost", "bNGFD", "SCFpre", "SCFpost", "SCFD",
              "SCGFpre", "SCGFpost", "SCGFD", "SDF1apre", "SDF1apost", "SDF1aD",
              "TNFbpre", "TNFbpost", "TNFbD", "TRAILpre", "TRAILpost", "TRAILD",
              "IL1bpre", "IL1bpost", "IL1bD", "IL1rapre", "IL1rapost", "IL1raD",
              "IL2pre", "IL2post", "IL2D", "IL4pre", "IL4post", "IL4D", "IL5pre",
              "IL5post", "IL5D", "IL6pre", "IL6post", "IL6D", "IL7pre", "IL7post",
              "IL7D", "IL8pre", "IL8post", "IL8D", "IL9pre", "IL9post", "IL9D",
              "IL10pre", "IL10post", "IL10D", "IL12pre", "IL12post", "IL12D",
              "IL13pre", "IL13post", "IL13D", "IL15pre", "IL15post", "IL15D",
              "IL17pre", "IL17post", "IL17D", "EOTpre", "EOTpost", "EOTD",
              "FGFbpre", "FGFbpost", "FGFbD", "GCSFpre", "GCSFpost", "GCSFD",
              "GMCSFpre", "GMCSFpost", "GMCSFD", "IFNgpre", "IFNgpost", "IFNgD",
              "IP10pre", "IP10post", "IP10D", "MCP1pre", "MCP1post", "MCP1D",
              "MIP1apre", "MIP1apost", "MIP1aD", "MIP1bpre", "MIP1bpost", "MIP1bD",
              "PDGFpre", "PDGFpost", "PDGFD", "RANTESpre", "RANTESpost", "RANTESD",
              "TNFapre", "TNFapost", "TNFaD", "VEGFpre", "VEGFpost", "VEGFD")

```

## The Data

The outcome is labelled `PPS` (0 or 1). Variables are named with
suffixes `pre` and `post` and the difference (`post` - `pre`) is
labelled with a suffix `D`. The cytokine data starts at column 27 (GT: I removed the RACHS1-Score, because it is not complete and will dampen the cytokine predictiveness. COuld compare results with and wothout RACHS1 later.)

#GT: I removed all Cop and ADM values because they stem from a separate analysis and do not qualify as cytokines!

```{r}
naRows <- c(4:5, 7:8)
d <- d[-naRows, ]
colNames <- names(d)
preBioNames <- grep("pre$", colNames[27:ncol(d)], value = TRUE)
bioNames <- gsub("pre$", "", preBioNames)
preBioData <- d[, preBioNames]
postBioNames <- grep("post$", colNames[27:ncol(d)], value = TRUE)
postBioData <- d[, postBioNames]
difNames <- grep("D$", colNames[27:ncol(d)], value = TRUE)
difData <- d[, difNames]
nonBioNames <- setdiff(colNames, c(preBioNames, postBioNames, difNames))
```

## The Scientific Questions

1. What variables are predictive of the `PPS` outcome pre-operatively?
2. What variables are associated with the `PPS` outcome, using data
both pre and post operation?
3. How good is the prediction? How does it compare with the AUC
   characteristics computed by the biostatistical firm hired by
   Dr. Theilmeier?

The following variables are known pre, and post operatively.

```{r}
preOpVars <- c("gender", "age", "cyanosis", "expECCtime",
               "expXclamp", "bodyweight")
postOpVars <- c("procedure", "orgfail", "ARF", "circfail", "lungfail",
                "ventilator", "ICU", "hospital", "mortality", "timeOnCPB",
                "xclamp", "circarrest", "fluidbal", "relfluidbal",
                "catecholamine", "infection", "antibiotics", "posBC")
```

## Data Transformations

We plot a few random biomarkers on the natural scale as well as the
log scale to decide what transformation to use.

```{r}
makeBothPlots <- function(bioName) {
    d[, bioName] %>%
        unlist ->
        value
    p1 <- ggplot() + geom_histogram(mapping  = aes(x = value), fill = "brown") +
        labs(x = bioName)
    p2 <- ggplot() + geom_histogram(mapping  = aes(x = log(value + 1)), fill = "brown") +
        labs(x = sprintf("log(%s)", bioName))
    list(p1, p2)
}
```

### Pre Op Biomarker values

We look at the pre-op biomarkers.

```{r}
set.seed(16272)
bms <- preBioNames[sample(seq_along(preBioNames), 5)]
library(cowplot)

```

```{r}
plot_grid(plotlist = makeBothPlots(bms[1]), ncol = 2)
```


```{r}
plot_grid(plotlist = makeBothPlots(bms[2]), ncol = 2)
```


```{r}
plot_grid(plotlist = makeBothPlots(bms[3]), ncol = 2)
```


```{r}
plot_grid(plotlist = makeBothPlots(bms[4]), ncol = 2)
```


```{r}
plot_grid(plotlist = makeBothPlots(bms[5]), ncol = 2)
```


### Difference in post/pre

We look at the differences now. 

```{r}
set.seed(16272)
bms <- difNames[sample(seq_along(difNames), 5)]
```


```{r}
plot_grid(plotlist = makeBothPlots(bms[1]), ncol = 2)
```


```{r}
plot_grid(plotlist = makeBothPlots(bms[2]), ncol = 2)
```


```{r}
plot_grid(plotlist = makeBothPlots(bms[3]), ncol = 2)
```


```{r}
plot_grid(plotlist = makeBothPlots(bms[4]), ncol = 2)
```


```{r}
plot_grid(plotlist = makeBothPlots(bms[5]), ncol = 2)
```

### Using Logged biomarker data henceforth

Based on the plots above, I transformed all the biomarker data to a
log scale. It doesn't seem to make things worse!

Specifically, `x` was transformed to `log(x + 1)` where `x` was either
the pre or post data. For post-pre differences, I considered the
difference of the logs.

```{r}
logPreBio <- apply(d[, preBioNames],
                   2,
                   function(x) log(x + 1))
colnames(logPreBio) <- paste("Log", preBioNames, sep=".")
logPostBio <- apply(d[, postBioNames],
                    2,
                    function(x) log(x + 1))
colnames(logPostBio) <- paste("Log", postBioNames, sep=".")
diff <- logPostBio - logPreBio
colnames(diff) <- bioNames
d <- data.frame(d, logPreBio, logPostBio, diff)
d$response <- d$PPS
d$PPS <- factor(d$PPS)
```

## A PCA plot of Pre-Op Biomarker Data.

We look at the principal components.

```{r}
pcaCharts <- function(x) {
    x.var <- x$sdev^2
    x.pvar <- x.var/sum(x.var)
    print("proportions of variance:")
    d <- data.frame(pc = seq_along(x.var),
                    prop = x.pvar,
                    cum.prop. = cumsum(x.pvar))
    # print(kable(d))
  print(d)

    par(mfrow=c(2,2))
    plot(x.pvar,xlab="Principal component", ylab="Proportion of variance explained", ylim=c(0,1), type='b')
    plot(cumsum(x.pvar),xlab="Principal component", ylab="Cumulative Proportion of variance explained", ylim=c(0,1), type='b')
    screeplot(x)
    screeplot(x,type="l")
    par(mfrow=c(1,1))
}
```

```{r}
pc <- prcomp(logPreBio, scale = TRUE)
pcaCharts(pc)
```


```{r}
library(ggfortify)
df <- d[, preBioNames]
autoplot(prcomp(df), data = d, colour = 'PPS')
```

### Conclusion

_There is an odd pattern here._ Several of the non-PPS patients are on
a straight line, but there is still no clear separation. There is no
clear boundary though.


## Logistic Regression on the Pre-OP variables and biomarkers

```{r}
library(glmnet)
varsIncluded <- c("age", "gender", "bodyweight", "cyanosis", "expECCtime", "expXclamp")
varsIncludedpost<- c("age", "gender", "bodyweight", "cyanosis", "timeOnCPB",
              "xclamp", "circarrest")
# I am introducing the real POST-parameters for the analysis of the differences here, that we can use to construct the model pre/post! Am I doing this right? I also removed a double of "cyanosis" that had creeped into the list.

xData <- as.matrix(cbind(d[, varsIncluded], logPreBio))
response <- d$response
print(cbind(1:ncol(xData), colnames(xData)))
```


Fit the model.

```{r}
fit <- glmnet(x = xData, y = response, family = "binomial")
plot(fit, xvar = "dev", label = TRUE)
```

Cross validation.

```{r}
cvfit <- cv.glmnet(xData, y = response, family = "binomial", type.measure = "class")
plot(cvfit)
```

Print the $\lambda$ corresponding to the CV fit.

```{r}
cat("Minimum", cvfit$lambda.min, "\n")
cat("Lambda 1se", cvfit$lambda.1se, "\n")
```

So what are the features corresponding to the best $\lambda_{1se}$ and
$\lambda_{1se}$ fits?

```{r}
opt.lam <- c(cvfit$lambda.min, cvfit$lambda.1se)
coef(cvfit, s = opt.lam)
```

Let us use AUC as the measure of fit.  We have to use `nfolds = 5` to
account for the small sample sizes in cross validation.


```{r}
set.seed(8337)
cvfit <- cv.glmnet(xData, y = response, family = "binomial", type.measure = "auc", nfolds = 5)
plot(cvfit)
```

Print the $\lambda$ corresponding to the CV fit.

```{r}
cat("Minimum", cvfit$lambda.min, "\n")
cat("Lambda 1se", cvfit$lambda.1se, "\n")
```

Let us see what variables are given by these two fits.

```{r}
opt.lam <- c(cvfit$lambda.min, cvfit$lambda.1se)
coef(cvfit, s = opt.lam)
```

## Always including some non Biomarker variables in the model

```{r}
set.seed(8122337)
cvfit <- cv.glmnet(xData, y = response, family = "binomial", type.measure = "auc", nfolds = 5,
                   penalty.factor = c(rep(0, length(varsIncluded)),
                                      rep(1, nrow(xData) - length(varsIncluded))))

plot(cvfit)
```

Print the $\lambda$ corresponding to the CV fit.

```{r}
cat("Minimum", cvfit$lambda.min, "\n")
cat("Lambda 1se", cvfit$lambda.1se, "\n")
```

Let us see what variables are given by these two fits.

```{r}
opt.lam <- c(cvfit$lambda.min, cvfit$lambda.1se)
coef(cvfit, s = opt.lam)
```

I believe including all the non-biomarker variables is problematic as
it looks like many of them are correlated and they bump out many
biomarkers. In order to be conservative, it is probably sufficient to
keep just `gender` and `RACHS` score (which takes `age` and
`bodyweight` into account, if I remember correctly.)

### Conclusion

Unfortunately, there any signal in the biomarker data seems to be
correlated with non-biomarker variables to a high degree. Consider for
example, `bodyweight`. Here is the correlation table of `bodyweight`
against the pre-op biomarkers.



```{r}
bdywt.cor <- tibble(biomarker = preBioNames,
                    corr = sapply(preBioNames, function(x) cor(d[, x], d[, "bodyweight"])))
kable(bdywt.cor %>% arrange(corr))
```



The biomarkers are not adding much beyond these variables.

## Logistic Regression on the difference


```{r}
library(glmnet)
xData <- as.matrix(cbind(d[, varsIncludedpost], diff))
response <- d$response
print(cbind(1:ncol(xData), colnames(xData)))
```

Fit the model.

```{r}
fit <- glmnet(x = xData, y = response, family = "binomial")
plot(fit, xvar = "dev", label = TRUE)
```

Cross validation.

```{r}
cvfit <- cv.glmnet(xData, y = response, family = "binomial", type.measure = "class")
plot(cvfit)
```

Print the $\lambda$ corresponding to the CV fit.

```{r}
cat("Minimum", cvfit$lambda.min, "\n")
cat("Lambda 1se", cvfit$lambda.1se, "\n")
```

So what are the features corresponding to the best $\lambda_{1se}$ and
$\lambda_{1se}$ fits?

```{r}
opt.lam <- c(cvfit$lambda.min, cvfit$lambda.1se)
coef(cvfit, s = opt.lam)
```

Let us use AUC as the measure of fit.  We have to use `nfolds = 5` to
account for the small sample sizes in cross validation.


```{r}
set.seed(8337)
cvfit <- cv.glmnet(xData, y = response, family = "binomial", type.measure = "auc", nfolds = 5)
plot(cvfit)

Print the $\lambda$ corresponding to the CV fit.

```{r}
cat("Minimum", cvfit$lambda.min, "\n")
cat("Lambda 1se", cvfit$lambda.1se, "\n")
```

Let us see what variables are given by these two fits.

```{r}
opt.lam <- c(cvfit$lambda.min, cvfit$lambda.1se)
coef(cvfit, s = opt.lam)
```


## Always including some non Biomarker variables in the model

```{r}
set.seed(8122337)
cvfit <- cv.glmnet(xData, y = response, family = "binomial", type.measure = "auc", nfolds = 5,
                   penalty.factor = c(rep(0, length(varsIncludedpost)),
                                      rep(1, nrow(xData) - length(varsIncludedpost))))

plot(cvfit)
```

Print the $\lambda$ corresponding to the CV fit.

```{r}
cat("Minimum", cvfit$lambda.min, "\n")
cat("Lambda 1se", cvfit$lambda.1se, "\n")
```

Let us see what variables are given by these two fits.

```{r}
opt.lam <- c(cvfit$lambda.min, cvfit$lambda.1se)
coef(cvfit, s = opt.lam)
```

## Conclusion

Once again there is not much of a signal here from the biomarkers.


## Session Info
```{r}

sessionInfo()
```
